### 2020-11-18 latex/vlp.tex
Updated \mathbb{S}^1 to \Sone, made Gamma \simeq \SxR instead of eq.

\section{Dynamics of the Variable Length Pendulum}
% Derive the dynamics of the VLP in Hamiltonian form
We will model the VLP as a point mass \(m\)
connected to a fixed pivot by a massless rod of varying length \(l\) with angle 
\(q \in \mathbb{S}^1\) from the vertical, as in Figure
\ref{fig:vlp-model}. 
We will ignore any damping and frictional forces in this model.
In a realistic VLP, the rod length \(l\) varies between some minimum
length \(\underline{l} \geq 0\) and some maximum length 
\(\overline{l} > \underline{l}\). The configuration of the VLP is the vector
\(\mathbf{q} := (q,l) \in \mathbb{S}^1 \times [\underline{l},\overline{l}]\).

@@

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The Cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos(q)\dot{q}, -\dot{l}\cos(q) + l\sin(q)\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
   .
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
   .
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
   .
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
   .
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we get the Hamiltonian ~\eqref{eqn:simply-actuated-hamiltonian} whose
dynamics ~\eqref{eqn:simply-actuated-dynamics} resolve to  
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q)
      , \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau
        . \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \mathbb{R}\) affecting the dynamics of
\(p_l\), acting collinearly with the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This makes the VLP into an underactuated mechanical system with degree of
underactuation one. 
It is also a useful assumption because it means \((\mathbf{q},\mathbf{p})\) 
are simply actuated coordinates, which allows us to apply the theory of VNHCs we
developed in Chapter \ref{ch:vnhcs}.

Let us define the VNHC \(l = L(q,p)\), by which we mean we are
actually defining the VNHC \(h(\mathbf{q},\mathbf{p}) = l - L(q,p) = 0\) of
order 1.
The VLP satisfies
\(\nabla_q \Minv(\mathbf{q}) = \Zmat{2 \times 2}\).
By Theorem \ref{thm:vnhc-regularity},
\(h(\mathbf{q},\mathbf{p})\) is a regular VNHC whenever 
\(L(q,p)\) is \(C^2\) because
\[
   dh_{\mathbf{q}} \Minv(\mathbf{q})B = 
   \begin{bmatrix}
      -\pdiff{L}{q} & 1
   \end{bmatrix}
   \begin{bmatrix}
      \frac{1}{ml^2}  & 0 \\
      0 & \frac{1}{m}
   \end{bmatrix} 
   \begin{bmatrix}
      0 \\ 
      1
   \end{bmatrix}
   = \frac{1}{m}
\]
is always full rank.
The constraint manifold is
\(\Gamma = \mathbb{S}^1 \times \R\), which is parameterized by the unactuated
phase \((q,p)\).
By Theorem \ref{thm:zero-dynamics}, the constrained dynamics
are described entirely by \((\dot{q},\dot{p})\) with \(l\)
replaced by \(L(q,p)\):
\begin{equation}\label{eqn:vlp-hamiltonian}
   \begin{cases}
      \dot{q} = \frac{p}{m L^2} \\
      \dot{p} = -mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
Note that we suppress the function notation of \(L(q,p)\) for clarity.

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \mathbb{S}^1 \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) for the VLP injects energy 
   on the constraint manifold \(\Gamma = \mathbb{S}^1 \times \R\) if there
   exists \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
   \end{equation}
   with the property that the inequality is strict except at 
   \(\theta \in \{0, \frac{\pi}{2}, \frac{\pi}, \frac{3\pi}{2}\}\).  
   If instead 
   \[
    \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \leq 0 
         \text{ }\forall \theta \in \mathbb{S}^1
         ,
   \]
   the VNHC dissipates energy on \(\Gamma\).
\end{thm}
\begin{proof}
We will first show that, under the given assumptions, the origin of the
constrained dynamics~\eqref{eqn:vlp-hamiltonian} is a repeller, and for that
we consider the negative time system obtained
from~\eqref{eqn:vlp-hamiltonian} by reversing the sign of the vector field:
\begin{equation}\label{eqn:negative-time-vlp}
   \begin{cases}
      \dot{q} = -\frac{p}{m L^2} \\
      \dot{p} = mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
The state space of~\eqref{eqn:negative-time-vlp} is 
\(\mathbb{S}^1 \times \R\), and the system has two equilibria,
\((q,p) = (0,0)\) and \((q,p) = (\pi,0)\). 
We need to show that the equilibrium \((q,p) = (0,0)\) is
asymptotically stable for~\eqref{eqn:negative-time-vlp}.

Consider the Lyapunov function candidate
\[
    E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
        + m g l_\text{avg} (1-\cos(q))
   ,
\]
which corresponds to the energy of VLP when the pendulum length is fixed at
\(l_\text{avg}\). 
The function \(E_\text{avg}\) is positive definite at \((0,0)\), and has compact
sublevel sets on \(\mathbb{S}^1 \times \R\).  
The derivative of \(E_\text{avg}\) along~\eqref{eqn:negative-time-vlp} is
\begin{equation}\label{eqn:eavg-dot-neg}
   \dot{E}_\text{avg} = \frac{g\sin(q)p \left(L(\theta)^3 -
     l_\text{avg}^3\right)} {l_\text{avg}^2L(\theta)^2}
   .
\end{equation}
It is easy to show that
\[
   \sign{\sin(q)p} = \sign{\sin(2\theta)}
   .
\]
Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
\[ 
    \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
    ,
\]
and therefore,
\begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}} &= 
       \sign{\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
    \nonumber \\
    &= \sign{\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)}
    \nonumber \\
    &= -\sign{(l_\text{avg} - L(\theta))\sin(2\theta)} 
    \nonumber \\
    &\leq 0 \text{ (by assumption)}
   .
\end{align}
We have thus shown that \(\dot{E}_\text{avg} \leq 0\), and therefore
the equilibrium \((q,p)=(0,0)\) is stable for~\eqref{eqn:negative-time-vlp}.
We now apply the Krasovskii-LaSalle invariance principle, and consider the
largest subset of 
\( Z = \left\{(q,p) \in \mathbb{S}^1 \times \R \mid \dot{E}_\text{avg}(q,p) = 0
\right\} \). 
We see from~\eqref{eqn:eavg-dot-neg}
that \(\dot{E}_\text{avg}(q,p)\) is zero when \(p = 0\), \(\sin(q) = 0\), or
\(L(\theta) = l_\text{avg}\). 
This latter condition by assumption is met when 
\(\theta \in \{0,\pi/2,\pi, 3\pi/2\}\), or equivalently when either 
\(q = 0\) or \(p = 0\). 
All in all, we have that
\[
   Z= \left\{(q,p) \in \mathbb{S}^1 \times \R \mid q =0\right\} 
   \cup \left\{(q,p) \in \mathbb{S}^1 \times \R \mid p =0\right\} 
   .
\]
It is easily seen that the largest invariant subset of \(Z\) is the union of the
two equilibria
\[
   \Omega =  \{(0,0)\}\cup \{(\pi,0)\}
   .
\]  
Since all sublevel sets of \(E_\text{avg}\) are compact, the
Krasovskii-LaSalle invariance principle implies that the set
\(\Omega\) is globally attractive \cite{krasovskii_lasalle}. 
Since the two equilibria are isolated, and since \((q,p)=(0,0)\) is stable, we
deduce that the equilibrium \((q,p)=(0,0)\) is asymptotically stable for the
negative time system~\eqref{eqn:negative-time-vlp}, and thus it is a repeller
for the constrained dynamics in~\eqref{eqn:vlp-hamiltonian}.

We now prove that almost all solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} escape compact sets in finite
time, \ie, that for each compact subset \(K\) of 
\(\mathbb{S}^1 \times \R\), and for almost every condition 
\( (q_0,p_0) \in K \), there exists \( T > 0\) such that
\( (q(t),p(t)) \notin K \) for all \( t > T \), where \( (q(t),p(t)) \)
denotes the solution of~\eqref{eqn:vlp-hamiltonian} with 
\( (q(0),p(0)) = (q_0,p_0)\). 
In what follows, we denote by \( x = (q,p) \) the state of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}. 
Also, we denote by \(\Pi^-\) the stable manifold of the equilibrium \((\pi,0)\)
for system~\eqref{eqn:vlp-hamiltonian}, defined as
\[
   \Pi^- := \left\{ x(0) \in \mathbb{S}^1 \times \R \mid
   \lim_{t \to \infty} x(t) = (\pi, 0) \right\}.
\]
The linearization of system~\eqref{eqn:vlp-hamiltonian} at the
equilibrium \((\pi,0)\) has a system matrix given by
\begin{equation}\label{eqn:linearized-vlp}
   \begin{bmatrix}
      0 & -\frac{1}{mL(0)^2} \\
      -m g L(0) & 0 \\
   \end{bmatrix}
   ,
\end{equation}
whose spectrum is \(\{\pm \sqrt{{g}/{L(0)}}\}\).  
Since one eigenvalue of ~\eqref{eqn:linearized-vlp} is positive and one is
negative, the stable manifold theorem implies that the stable manifold 
\( \Pi^- \) is a one-dimensional immersed submanifold of
\(\mathbb{S}^1 \times \R\), and thus is a set of measure zero
\cite{stable_manifold}.

Let \( K \subset \mathbb{S}^1 \times \R\) be an arbitrary compact set and let 
\( x(0) \in K \backslash \left(\Pi^- \cup \left\{(0,0)\right\}\right)\).
Since \( x(0) \notin \Pi^- \), we have that
\begin{equation}\label{eqn:vlp-x-does-not-converge}
   x(t) \nrightarrow_{t \to \infty} (\pi,0).
\end{equation}
Suppose, by way of contradiction, that for each \(T > 0\) there exists 
\(t' > T\) such that \(x(t') \in K\).  
Letting
\[
   k := \max\limits_{x \in K} E_\text{avg}(x)
   ,
\] 
and \(E_k =\{ x \mid E_\text{avg}(x) \leq k\} \), we have that 
\( K \subset E_k \), and thus \(x(t') \in E_k\).
We have shown earlier that the function \( E_\text{avg} \) is nonincreasing
along solutions of~\eqref{eqn:negative-time-vlp}, which implies that
\( E_\text{avg}(x(t)) \) is nondecreasing for solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}, and therefore the half orbit \(x([0,t'])\)
is contained fully within \(E_k\).  
In particular, \(x(T) \in E_k\). Since this is true for each \( T >0 \), we
deduce that \(x(t) \in E_k\) for all \(t \in [0,\infty[\).  
Since \(E_k\) is compact, \(x(t)\) has a positive limit set in \(E_k\) which, by
the Poincar\'{e}-Bendixson theorem, is either an equilibrium or a closed orbit
\cite{poincare_bendixson}.  
We will show this positive limit set must be an equilibrium and thereby reach a
contradiction.

@@

We conclude that the VNHC injects energy into the VLP on 
\(\mathbb{S}^1 \times \R\).
By flipping the inequality of ~\eqref{eqn:vlp-energy-gain-condition}
we find the VLP is gaining energy in negative-time, so the VNHC is
dissipating energy.
\end{proof}

@@

>>>>

\section{Dynamics of the Variable Length Pendulum}
% Derive the dynamics of the VLP in Hamiltonian form
We will model the VLP as a point mass \(m\)
connected to a fixed pivot by a massless rod of varying length \(l\) with angle 
\(q \in \Sone\) from the vertical, as in Figure
\ref{fig:vlp-model}. 
We will ignore any damping and frictional forces in this model.
In a realistic VLP, the rod length \(l\) varies between some minimum
length \(\underline{l} \geq 0\) and some maximum length 
\(\overline{l} > \underline{l}\). The configuration of the VLP is the vector
\(\mathbf{q} := (q,l) \in \Sone \times [\underline{l},\overline{l}]\).

@@

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The Cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos(q)\dot{q}, -\dot{l}\cos(q) + l\sin(q)\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
   .
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
   .
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
   .
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
   .
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we get the Hamiltonian ~\eqref{eqn:simply-actuated-hamiltonian} whose
dynamics ~\eqref{eqn:simply-actuated-dynamics} resolve to  
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q)
      , \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau
        . \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \R\) affecting the dynamics of
\(p_l\), acting collinearly with the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This makes the VLP into an underactuated mechanical system with degree of
underactuation one. 
It is also a useful assumption because it means \((\mathbf{q},\mathbf{p})\) 
are simply actuated coordinates, which allows us to apply the theory of VNHCs we
developed in Chapter \ref{ch:vnhcs}.

Let us define the VNHC \(l = L(q,p)\), by which we mean we are
actually defining the VNHC \(h(\mathbf{q},\mathbf{p}) = l - L(q,p) = 0\) of
order 1.
The VLP satisfies
\(\nabla_q \Minv(\mathbf{q}) = \Zmat{2 \times 2}\).
By Theorem \ref{thm:vnhc-regularity},
\(h(\mathbf{q},\mathbf{p})\) is a regular VNHC whenever 
\(L(q,p)\) is \(C^2\) because
\[
   dh_{\mathbf{q}} \Minv(\mathbf{q})B = 
   \begin{bmatrix}
      -\pdiff{L}{q} & 1
   \end{bmatrix}
   \begin{bmatrix}
      \frac{1}{ml^2}  & 0 \\
      0 & \frac{1}{m}
   \end{bmatrix} 
   \begin{bmatrix}
      0 \\ 
      1
   \end{bmatrix}
   = \frac{1}{m}
\]
is always full rank.
The constraint manifold \(\Gamma\) is diffeomorphic to 
\(\SxR\), which is parameterized by the unactuated
phase \((q,p)\).
By Theorem \ref{thm:zero-dynamics}, the constrained dynamics
are described entirely by \((\dot{q},\dot{p})\) with \(l\)
replaced by \(L(q,p)\):
\begin{equation}\label{eqn:vlp-hamiltonian}
   \begin{cases}
      \dot{q} = \frac{p}{m L^2} \\
      \dot{p} = -mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
Note that we suppress the function notation of \(L(q,p)\) for clarity.

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \Sone \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) for the VLP injects energy 
   on the constraint manifold \(\Gamma \simeq \SxR\) if there
   exists \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \Sone
      ,
   \end{equation}
   with the property that the inequality is strict except at 
   \(\theta \in \{0, \frac{\pi}{2}, \frac{\pi}, \frac{3\pi}{2}\}\).  
   If instead 
   \[
    \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \leq 0 
         \text{ }\forall \theta \in \Sone
         ,
   \]
   the VNHC dissipates energy on \(\Gamma\).
\end{thm}
\begin{proof}
We will first show that, under the given assumptions, the origin of the
constrained dynamics~\eqref{eqn:vlp-hamiltonian} is a repeller, and for that
we consider the negative time system obtained
from~\eqref{eqn:vlp-hamiltonian} by reversing the sign of the vector field:
\begin{equation}\label{eqn:negative-time-vlp}
   \begin{cases}
      \dot{q} = -\frac{p}{m L^2} \\
      \dot{p} = mgL\sin(q)
      . \\ 
   \end{cases}
\end{equation}
The state space of~\eqref{eqn:negative-time-vlp} is 
\(\SxR\), and the system has two equilibria,
\((q,p) = (0,0)\) and \((q,p) = (\pi,0)\). 
We need to show that the equilibrium \((q,p) = (0,0)\) is
asymptotically stable for~\eqref{eqn:negative-time-vlp}.

Consider the Lyapunov function candidate
\[
    E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
        + m g l_\text{avg} (1-\cos(q))
   ,
\]
which corresponds to the energy of VLP when the pendulum length is fixed at
\(l_\text{avg}\). 
The function \(E_\text{avg}\) is positive definite at \((0,0)\), and has compact
sublevel sets on \(\SxR\).  
The derivative of \(E_\text{avg}\) along~\eqref{eqn:negative-time-vlp} is
\begin{equation}\label{eqn:eavg-dot-neg}
   \dot{E}_\text{avg} = \frac{g\sin(q)p \left(L(\theta)^3 -
     l_\text{avg}^3\right)} {l_\text{avg}^2L(\theta)^2}
   .
\end{equation}
It is easy to show that
\[
   \sign{\sin(q)p} = \sign{\sin(2\theta)}
   .
\]
Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
\[ 
    \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
    ,
\]
and therefore,
\begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}} &= 
       \sign{\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
    \nonumber \\
    &= \sign{\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)}
    \nonumber \\
    &= -\sign{(l_\text{avg} - L(\theta))\sin(2\theta)} 
    \nonumber \\
    &\leq 0 \text{ (by assumption)}
   .
\end{align}
We have thus shown that \(\dot{E}_\text{avg} \leq 0\), and therefore
the equilibrium \((q,p)=(0,0)\) is stable for~\eqref{eqn:negative-time-vlp}.
We now apply the Krasovskii-LaSalle invariance principle, and consider the
largest subset of 
\( Z = \left\{(q,p) \in \SxR \mid \dot{E}_\text{avg}(q,p) = 0
\right\} \). 
We see from~\eqref{eqn:eavg-dot-neg}
that \(\dot{E}_\text{avg}(q,p)\) is zero when \(p = 0\), \(\sin(q) = 0\), or
\(L(\theta) = l_\text{avg}\). 
This latter condition by assumption is met when 
\(\theta \in \{0,\pi/2,\pi, 3\pi/2\}\), or equivalently when either 
\(q = 0\) or \(p = 0\). 
All in all, we have that
\[
   Z= \left\{(q,p) \in \SxR \mid q =0\right\} 
   \cup \left\{(q,p) \in \SxR \mid p =0\right\} 
   .
\]
It is easily seen that the largest invariant subset of \(Z\) is the union of the
two equilibria
\[
   \Omega =  \{(0,0)\}\cup \{(\pi,0)\}
   .
\]  
Since all sublevel sets of \(E_\text{avg}\) are compact, the
Krasovskii-LaSalle invariance principle implies that the set
\(\Omega\) is globally attractive \cite{krasovskii_lasalle}. 
Since the two equilibria are isolated, and since \((q,p)=(0,0)\) is stable, we
deduce that the equilibrium \((q,p)=(0,0)\) is asymptotically stable for the
negative time system~\eqref{eqn:negative-time-vlp}, and thus it is a repeller
for the constrained dynamics in~\eqref{eqn:vlp-hamiltonian}.

We now prove that almost all solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian} escape compact sets in finite
time, \ie, that for each compact subset \(K\) of 
\(\SxR\), and for almost every condition 
\( (q_0,p_0) \in K \), there exists \( T > 0\) such that
\( (q(t),p(t)) \notin K \) for all \( t > T \), where \( (q(t),p(t)) \)
denotes the solution of~\eqref{eqn:vlp-hamiltonian} with 
\( (q(0),p(0)) = (q_0,p_0)\). 
In what follows, we denote by \( x = (q,p) \) the state of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}. 
Also, we denote by \(\Pi^-\) the stable manifold of the equilibrium \((\pi,0)\)
for system~\eqref{eqn:vlp-hamiltonian}, defined as
\[
   \Pi^- := \left\{ x(0) \in \SxR \mid
   \lim_{t \to \infty} x(t) = (\pi, 0) \right\}.
\]
The linearization of system~\eqref{eqn:vlp-hamiltonian} at the
equilibrium \((\pi,0)\) has a system matrix given by
\begin{equation}\label{eqn:linearized-vlp}
   \begin{bmatrix}
      0 & -\frac{1}{mL(0)^2} \\
      -m g L(0) & 0 \\
   \end{bmatrix}
   ,
\end{equation}
whose spectrum is \(\{\pm \sqrt{{g}/{L(0)}}\}\).  
Since one eigenvalue of ~\eqref{eqn:linearized-vlp} is positive and one is
negative, the stable manifold theorem implies that the stable manifold 
\( \Pi^- \) is a one-dimensional immersed submanifold of
\(\SxR\), and thus is a set of measure zero
\cite{stable_manifold}.

Let \( K \subset \SxR\) be an arbitrary compact set and let 
\( x(0) \in K \backslash \left(\Pi^- \cup \left\{(0,0)\right\}\right)\).
Since \( x(0) \notin \Pi^- \), we have that
\begin{equation}\label{eqn:vlp-x-does-not-converge}
   x(t) \nrightarrow_{t \to \infty} (\pi,0).
\end{equation}
Suppose, by way of contradiction, that for each \(T > 0\) there exists 
\(t' > T\) such that \(x(t') \in K\).  
Letting
\[
   k := \max\limits_{x \in K} E_\text{avg}(x)
   ,
\] 
and \(E_k =\{ x \mid E_\text{avg}(x) \leq k\} \), we have that 
\( K \subset E_k \), and thus \(x(t') \in E_k\).
We have shown earlier that the function \( E_\text{avg} \) is nonincreasing
along solutions of~\eqref{eqn:negative-time-vlp}, which implies that
\( E_\text{avg}(x(t)) \) is nondecreasing for solutions of the reduced
dynamics~\eqref{eqn:vlp-hamiltonian}, and therefore the half orbit \(x([0,t'])\)
is contained fully within \(E_k\).  
In particular, \(x(T) \in E_k\). Since this is true for each \( T >0 \), we
deduce that \(x(t) \in E_k\) for all \(t \in [0,\infty[\).  
Since \(E_k\) is compact, \(x(t)\) has a positive limit set in \(E_k\) which, by
the Poincar\'{e}-Bendixson theorem, is either an equilibrium or a closed orbit
\cite{poincare_bendixson}.  
We will show this positive limit set must be an equilibrium and thereby reach a
contradiction.

@@

We conclude that the VNHC injects energy into the VLP on 
\(\SxR\).
By flipping the inequality of ~\eqref{eqn:vlp-energy-gain-condition}
we find the VLP is gaining energy in negative-time, so the VNHC is
dissipating energy.
\end{proof}

@@

% TODO: Add a remark:
% Our controller not only injects energy, but makes the lower equilibrium a
% repeller which means the "bad" set of x(0) does not contain a nbhd of (0,0),
% so anything near (0,0) will start to gain eenrgy aka ANY slight deviation from
% 0 will gain energy.
